Gap formation and stability in non-isothermal protoplanetary discs

### 2015-03-27 paper/summary.tex
backup

% {\bf question: vortex amplitude plot suggest just after linear growth,
%   vortex amplitudes are smaller with increasing beta, but later on in
%   the quasi-steady state, this trend reverses? yes. 
% }
\section{Summary and discussion}\label{summary}
In this paper, we have carried out numerical simulations of
non-isothermal disc-planet interaction.  
Our simulations were customized to examine the effect of a finite 
cooling time on the stability of gaps  
opened by giant planets to the so-called vortex or Rossby wave
instability. To do so, we  
included an energy equation with a cooling term that restores the 
disc temperature to its initial profile on a characteristic timescale
$t_c$. We studied the evolution of the gap stability as a function of 
$t_c$. This is a natural extension to previous studies of on gap
stability, which employ locally or strictly isothermal equations of
state. We considered the inviscid limit which favors the RWI
  \citep{li09,fu14} and avoids complications from viscous
  heating other that shock heating. However, this means that the vortex lifetimes observed in
  our simulations are likely longer than in realistic discs with
  non-zero physical viscosity.

@@

In the second set of calculations, we included the planet potential
throughout the simulations  and examined the long-term evolution of
the gap-edge vortex that develops from the RWI. The vortex reaches 
a quasi-steady state lasting $O(10^3)$ orbits. Unlike the `planet-off'
simulations, in which vortices decay after linear growth and merging,
we find that with the planet potential kept on, the vortex amplitude
grows during this quasi-steady state, during which no vortex migration is
  observed, until it begins to induce
shocks, after which the vortex amplitude begins to decay.

{\bf For our main simulations with $\tilde{\beta}\geq 0.1$}, the duration of the quasi-steady state increases with
increasing cooling timescales until a critical value, beyond which this
quasi-steady state shortens again. {\bf We find the timescale for the vortex to
decay, after reaching maximum amplitude, can be long for small
$\tilde{\beta}$ which contributes to a long overall vortex lifetime
with rapid cooling.} We do observe vortex migration during {\bf its}
decay, which may influence this decay timescale.

% The sum of
% these two timescales results in a double-peaked vortex lifetime as a
% function of cooling timescale.

@@

For short cooling timescales, the planet is able to open a
deeper gap which favours the RWI, leading to stronger 
vortices. For long cooling timescales, we find the vortex
grows faster during the quasi-steady state. In accordance with
previous stability calculations \citep{li00}, we 
suggest the latter is due to the RWI being favoured with increasing
disc temperature, and that this effect overcomes weaker
gap-opening for sufficiently long cooling times. 
These competing factors imply
for both short and long cooling timescales, the vortex reaches
its maximum amplitude, shock, and begins to decay, sooner than
intermediate cooling timescales. {(\bf However, for very rapid
  cooling, e.g. $\tilde{\beta}=0.01$, the quasi-steady state is also
  quite long. In this case other factors may be important, such as
  vortex migration.)}

@@

We remark that a non-monotonic dependence of the vortex lifetime was
also reported by \cite{fu14}, who performed isothermal disc-planet
simulations with different values of the   
disc aspect ratio. In their simulations the optimum aspect ratio is
$h=0.06$. In our simulations, $h$ is a dynamical
variable, but by analyzing the region where the vortex is located
($r-r_p\in[2,10]r_h$), % {\bf define this region}
we find for a dimensionless cooing timescale of $\tilde\beta=2.5$, which has
the longest vortex lifetime {\bf in the presence of moderate cooling}, that  
$h\approx0.058$ on average.  
%{\bf if new definition of vortex lifetime gives other $\tilde{\beta}$ values as the maximum lifetime, then need to update this value of $h$}. 
Our result is consistent with \cite{fu14}.

@@

  \emph{Orbital migration.} We have held the planet fixed on a circular
  orbit. However, gap-edge vortices are known to exert
  significant, oscillatory torques on the planet \citep{li09}
  which can lead to complex orbital migration. This will likely affect
  vortex lifetimes as it may alter the planet-vortex separation, as
  well as leading to direct vortex-planet interactions
  \citep{lin10,ataiee14}. Thus, 
  future simulations should allow the planet to freely
  migrate. Similarly, the role of vortex migration on its lifetime
  should be clarified.

  \emph{Cooling model.} Our prescription for the disc 
  heating/cooling is convenient to probe the full range of
  thermodynamic response of the disc. However, in order to calculate 
  vortex lifetimes in actual 
  protoplanetary discs, an improved thermodynamics treatment,
  e.g. radiative cooling based on realistic disc temperature, density,
  opacity models etc., should be used in future work.

>>>>

% {\bf question: vortex amplitude plot suggest just after linear growth,
%   vortex amplitudes are smaller with increasing beta, but later on in
%   the quasi-steady state, this trend reverses? yes. 
% }
\section{Summary and discussion}\label{summary}
In this paper, we have carried out numerical simulations of
non-isothermal disc-planet interaction.  
Our simulations were customized to examine the effect of a finite 
cooling time on the stability of gaps  
opened by giant planets to the so-called vortex or Rossby wave
instability. To do so, we  
included an energy equation with a cooling term that restores the 
disc temperature to its initial profile on a characteristic timescale
$t_c$. We studied the evolution of the gap stability as a function of 
$t_c$. This is a natural extension to previous studies of on gap
stability, which employ locally or strictly isothermal equations of
state. {\bf We considered the inviscid limit which favors the RWI
  \citep{li09,fu14} and avoids complications from viscous
  heating other that shock heating. However, this means that the vortex lifetimes observed in
  our simulations are likely longer than in realistic discs with
  non-zero physical viscosity.}

@@

In the second set of calculations, we included the planet potential
throughout the simulations  and examined the long-term evolution of
the gap-edge vortex that develops from the RWI. The vortex reaches 
a quasi-steady state lasting $O(10^3)$ orbits. Unlike the `planet-off'
simulations, in which vortices decay after linear growth and merging,
we find that with the planet potential kept on, the vortex amplitude
grows during this quasi-steady state{\bf, during which no vortex migration is
  observed}, until it begins to induce
shocks, after which the vortex amplitude begins to decay.

{\bf For our main simulations with $\tilde{\beta}\geq 0.1$}, the duration of the quasi-steady state increases with
increasing cooling timescales until a critical value, beyond which this
quasi-steady state shortens again. The timescale for the vortex to
decay, after reaching maximum amplitude, decreases monotonically with
increasing cooling time. {\bf We do observe vortex migration during this
  decay, which may influence this decay timescale. }

@@

For short cooling timescales, the planet is able to open a
deeper gap which favours the RWI, leading to stronger 
vortices. For long cooling timescales, we find the vortex
grows faster during the quasi-steady state. In accordance with
previous stability calculations \citep{li00}, we 
suggest the latter is due to the RWI being favoured with increasing
disc temperature, and that this effect overcomes weaker
gap-opening for sufficiently long cooling times. 
These competing factors imply
for both short and long cooling timescales, the vortex reaches
its maximum amplitude, shock, and begins to decay, sooner than
intermediate cooling timescales.

{(\bf However, for very rapid cooling, e.g. $\tilde{\beta}=0.01$, the quasi-steady state is also
  quite long. This suggests that the above effects themselves do not
  have a simple dependence on the cooling timescale when considering
  $\tilde{\beta}\to0$ and/or that other factors become important in
  this limit. This should be investigated in future works.)}

@@

We remark that a non-monotonic dependence of the vortex lifetime was
also reported by \cite{fu14}, who performed {\bf locally} isothermal disc-planet
simulations with different values of the   
disc aspect ratio. In their simulations the optimum aspect ratio is
$h=0.06$. In our simulations, $h$ is a dynamical
variable, but by analyzing the region where the vortex is located
($r-r_p\in[2,10]r_h$), % {\bf define this region}
we find for a dimensionless cooing timescale of $\tilde\beta=2.5$, which has
the longest vortex lifetime {\bf in the presence of moderate cooling}, that  
$h\approx0.058$ on average.  
%{\bf if new definition of vortex lifetime gives other $\tilde{\beta}$ values as the maximum lifetime, then need to update this value of $h$}. 
Our result is consistent with \cite{fu14}.

@@

{\bf
  \emph{Orbital migration.} We have held the planet fixed on a circular
  orbit. However, gap-edge vortices are known to exert
  significant, oscillatory torques on the planet \citep{li09}
  which can lead to complex orbital migration. This will likely affect
  vortex lifetimes as it may alter the planet-vortex separation, as
  well as leading to direct vortex-planet interactions
  \citep{lin10,ataiee14}. Thus, 
  future simulations should allow the planet to freely
  migrate. Similarly, the role of vortex migration on its lifetime
  should be clarified. 
}

{\bf 
  \emph{Cooling model.} Our prescription for the disc 
  heating/cooling is convenient to probe the full range of
  thermodynamic response of the disc. However, in order to calculate 
  vortex lifetimes in actual 
  protoplanetary discs, an improved thermodynamics treatment,
  e.g. radiative cooling based on realistic disc temperature, density,
  opacity models etc., should be used in future work.   
}
